Type theories as quotient inductive-recursive types

### 2025-02-06 tex/main.tex
IR background

\Fredrik{I'm not sure if I have included all points here. I tried to stay more positive re overlapping patterns.}
\LT{Thanks!}

@@

\cite{UFP2013}
\LT{
Points to include:
\begin{enumerate}
  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality~\cite{McBride1999} in \Agda $x \simeq y$ for $x : A$ and $y : B$
  \item We work with intensional type theory with uniqueness of identity proof and function extensionality (only used for NbE and the standard model for the extensional identity type).
  \item We use Agda to formally implement our definitions with the following options: \texttt{-{}-with-K}, \texttt{-{}-local-confluence-check}, \texttt{-{}-exact-split}, and \texttt{-{}-rewriting} using postulated equations~\cite{Licata2011} to introduce equality constructors for quotient types.
\end{enumerate}}

\LT{The symbol \textdbend indicates a property that is proved informally or checked partially with \Agda (because of postulated quotient inductive types).}
\subsection{Inductive-recursive types}
\cite{Dybjer2003,Dybjer2000,Dybjer1999}
\LT{Introduce abbreviations here}

\LT{Mention that the recursion part of induction-recursion are already not defined by elimination rule.}

@@

\section{The elimination principle and models}
\LT{Terminology: the eliminator principle vs.\ eliminator vs.\ induction principle?}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).

@@

\subsection{The equivalence} \label{subsec:equivalence}
\section{Formal implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Discussion}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}

>>>>

@@

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B(x)$.
%
We are further inspired by Agda in writing $\implicit{x}B(x)$ for $(x : A) \to B(x)$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.

We write $x \jdeq y$ for definitional equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the ``dependent'' identity type where $P : A \to \Set$, $t : P(x)$, $u : P(y)$ and $p : x =^A y$. One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$, where $\transfibf{P} : (p : x =^A y) \to P(x) \to P(y)$ is our notation for the operation which allows is to transport values between type families applied to elements that are equal up to the identity type.
%
Finally, we write $x \simeq y$ for the \emph{heterogeneous} identity type, where we allow $x : A$ and $y : B$ to have potentially different types.
As shown by McBride~\cite{McBride1999}, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.

We have formally implemented our definitions and proofs in Agda, with the following options: \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}. We simulate quotient inductive types using postulated equations to introduce equality constructors~\cite{Licata2011}.
Some of our theorem statements are meta-theoretical, and thus cannot be checked by Agda. Others can only be checked  partially, because of postulated quotient inductive types. We have marked such ``informal'' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.

%\LT{
%Points to include:
%\begin{enumerate}
%  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality~\cite{McBride1999} in \Agda $x \simeq y$ for $x : A$ and $y : B$
%  \item We work with intensional type theory with uniqueness of identity proof and function extensionality (only used for NbE and the standard model for the extensional identity type).
%  \item We use Agda to formally implement our definitions with the following options: \texttt{-{}-with-K}, \texttt{-{}-local-confluence-check}, \texttt{-{}-exact-split}, and \texttt{-{}-rewriting} using postulated equations~\cite{Licata2011} to introduce equality constructors for quotient types.
%\end{enumerate}}
%
%\LT{The symbol \textdbend indicates a property that is proved informally or checked partially with \Agda (because of postulated quotient inductive typeswbe checked ).}

\subsection{Inductive-recursive types}

An ordinary inductive type $A$, such as for example the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined by recursion, by providing an elimination rule which says that such a function $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$. For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\mathsf{suc}\,n)$, where the latter might refer to the already computed value $f(n)$.
%
The type is first defined inductively by listing its constructors, and then afterwards recursive functions can be defined out of it.

In contrast, an inductive-recursive type is defined inductively \emph{simultaneously} with a recursively defined function out of it. The prototypical example is Martin-L\"of's universe \`a la Tarski, where a type $U$ of codes for small types is defined simultaneously with a ``decoding'' function $T : U \to \Set$ which maps each code to the type it represents. Crucially, the code for dependent function types
\[
  \pi : (a : U) \to (T(a) \to U) \to U
\]
with $T(\pi(a, b)) = (x : T(a)) \to T(b(x))$ only makes sense because we already have access to the decoding function $T$.
%
Inductive-recursive types (IRT) were introduced by Dybjer~\cite{Dybjer2000}, and then extensively studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}.
The general schema allows one to define a type $A$ inductively, together with a recursive function $F : A \to D$, for some fixed (possibly large) type $D$.

An IRT $(A, F)$ comes with an elimination rule that allows to define a dependent function  $f : (x : A) \to P(x)$ by giving values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$, again reflecting that $A$ is inductively defined.
%
However, note that the recursive function $F : A \to D$ is not defined using this elimination rule, but part of the definition of the type itself.
%
In particular, the elimination rule can be used to define endofunctions on $A$, but the recursive function  $F : A \to D$ must always target a fixed pre-existing type $D$.
%
In this work, we will consider an extension \emph{generalised} IRTs, where also $F : A \to A$ is allowed.
%
Such IRTs were previously used by Danielsson~\cite{Danielsson2006} to represent type theory in type theory, and are discussed by Kaposi~\cite{Kaposi2023}.
%
Their general metatheory would be interesting future work.

%\LT{Mention that the recursion part of induction-recursion are already not defined by elimination rule.}

@@

\section{The elimination principle and models} \label{sec:models}
\LT{Terminology: the eliminator principle vs.\ eliminator vs.\ induction principle?}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).

@@

\subsection{The equivalence} \label{subsec:equivalence}
\section{Formal implementation}\label{sec:formal-implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Discussion}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}
